Simulations

### 2025-11-13 mutual_information_curves.qmd
Recommit changes without large RData file

mut_info <- function(sims, num_teams, rank_weights = NULL, wtype = "symmetric_salience"){
  # Handle missing weights (default to equal weights if not provided)
  if (is.null(rank_weights)){
    rank_weights <- rep(1,num_teams)
  }
  # Extract y values from sims, assuming `rank_hat` contains a list of rankings per simulation
  y_df <- sims %>%
    group_by(simulation) %>%
    summarise(rank_hat = list(rank_hat)) %>%  # Ensure rank_hat is a list column
    mutate(rank_hat = map(rank_hat, unlist)) %>%  # Unlist rank_hat properly
    unnest_wider(rank_hat, names_sep = "_")  # Expand into separate columns

  # Rename y columns
  colnames(y_df)[-1] <- paste0("y", 1:num_teams)

  # Create x matrix, representing positions 1:num_teams
  x_matrix <- matrix(rep(1:num_teams, times = nrow(y_df)), ncol = num_teams, byrow = TRUE)
  x_df <- as.data.frame(x_matrix)
  colnames(x_df) <- paste0("x", 1:num_teams)

  # Combine x and y data frames
  perms_df <- cbind(x_df, y_df %>% select(-simulation))
  perms_df <- perms_df %>%
    mutate(across(starts_with("x"), as.numeric)) %>%
    mutate(across(starts_with("y"), as.numeric))

  # Generate all possible permutations of rankings
  all_perms <- as.data.frame(do.call(rbind, permn(1:num_teams)))
  colnames(all_perms) <- paste0("y", 1:num_teams)

  # Count occurrences of each observed ranking
  prob_df <- perms_df %>%
    group_by(across(starts_with("x")),across(starts_with("y")))  %>%
    summarise(count = n(), .groups = "drop")

  # Merge with full permutation set to ensure all permutations exist
  prob_df <- right_join(all_perms, prob_df, by = colnames(all_perms)) %>%
    mutate(count = replace_na(count, 0)) %>%
    mutate(probability = count / sum(count))  # Normalize probabilities

@@

  for (i in 1:num_teams) {
    #x_i <- paste0("x", i)
    #y_i <- paste0("y", i)

    #max_footrule <- sum(rank_weights * abs((1:num_teams) - (num_teams:1)))
    # Loop through each row of prob_df
    for (j in 1:nrow(prob_df)) {
      x_vals <- as.numeric(prob_df[j, paste0("x", 1:num_teams)])
      y_vals <- as.numeric(prob_df[j, paste0("y", 1:num_teams)])

@@

    }
  }
  #w <- 1 / (w+1)  # Inverse of squared differences and manually change 0 probabilities to 1 on the inverse
  #w <- w/sum(w)
  # Compute weights dynamically for each row
  weighted_df <- prob_df %>%
    mutate(w = w)

@@

  Raj_dist <- 1-numerator/denominator
  # Return the weighted data frame with mutual information weights
  return(Raj_dist)
}

@@

mi_plots_df <- data.frame(
indices = 1:8,
"Single Elimination (Reseeded)" = 1-mi_se_reseeded,
"Group Stage" = 1-mi_group_stage,
"Double Elimination" = 1-mi_de_norm,
"Single Elimination (Set Structure)" = 1-mi_se_norm,
"Coin Flip Single Elimination" = 1-mi_random,
#"Unif_SE" = mi_se_unif,
"Bad Seeding Structure Single Elimination" = 1-mi_bad,
"Coin Flip Round Robin" = 1-mi_perms,
"Round Robin" = 1-mi_test_rr,
"Worst Possible" = 1-mi_worst,
"Best Possible" = 1-mi_best,
check.names = FALSE
)

@@

```{r}
compute_mut_plots_rydog <- function(dataset) {
  num_teams <- length(unique(dataset$true_rank))
  results <- numeric(num_teams)
  for (i in 1:num_teams) {
    binary_vector <- c(rep(1, i), rep(0, num_teams - i))  # Construct vector
    results[i] <- mut_info(dataset, num_teams, binary_vector, wtype = "rydog")  # Compute MI
  }
  return(results)
}

# Compute mutual information for each dataset
mi_se_reseeded_rydog <- compute_mut_plots_rydog(reseeded_results_eight)
mi_group_stage_rydog <- compute_mut_plots_rydog(group_stage_results_eight)
mi_de_norm_rydog <- compute_mut_plots_rydog(de_norm_results_eight)
mi_se_norm_rydog <- compute_mut_plots_rydog(all_results_norm)
mi_test_rr_rydog <- compute_mut_plots_rydog(test_rr_noTies)
mi_seven_se_rydog <- compute_mut_plots_rydog(seven_games_se_results)
mi_random_rydog <- compute_mut_plots_rydog(equal_strengths)
mi_se_unif_rydog <- compute_mut_plots_rydog(test_unif)
mi_bad_rydog <- compute_mut_plots_rydog(bad_seeding_structure)
mi_perms_rydog <- compute_mut_plots_rydog(perms_df)
mi_best_rydog <- compute_mut_plots_rydog(best)
mi_worst_rydog <- compute_mut_plots_rydog(worst)

mi_plots_df_rydog <- data.frame(
indices = 1:8,
"Single Elimination (Reseeded)" = 1-mi_se_reseeded_rydog,
"Group Stage" = 1-mi_group_stage_rydog,
"Double Elimination" = 1-mi_de_norm_rydog,
"Single Elimination (Set Structure)" = 1-mi_se_norm_rydog,
"Coin Flip Single Elimination" = 1-mi_random_rydog,
#"Unif_SE" = mi_se_unif,
"Bad Seeding Structure Single Elimination" = 1-mi_bad_rydog,
"Coin Flip Round Robin" = 1-mi_perms_rydog,
"Round Robin" = 1-mi_test_rr_rydog,
"Worst Possible" = 1-mi_worst_rydog,
"Best Possible" = 1-mi_best_rydog
)

# Reshape data into long format for ggplot
df_long_rydog <- pivot_longer(mi_plots_df_rydog, cols = -indices, names_to = "Method", values_to = "MI")
windowsFonts(Arial = windowsFont("Arial"))
ggplot(df_long_rydog, aes(x = indices, y = MI, color = Method, group = Method)) +
geom_line(linewidth = 2.5) +           # Add lines
geom_point(size=3) +          # Add points
labs(
title = "Comparison of Tournament Structures with Normal Strengths (Rydog)",
x = "Top i Number of Teams",
y = "Normalized Mutual Information \\ Entropy (log10 scale)"
) +
theme_minimal() +       # Use a clean theme
theme(text = element_text(family = "Arial", size = 32),
  legend.position = "bottom",
  legend.box = "vertical",
  plot.title = element_text(hjust = 0.5, size = 36),
  axis.title = element_text(size = 30),
  axis.text = element_text(size=30)) +
  guides(color = guide_legend(nrow = 5)) +
  scale_y_log10(
    labels = scales::label_number() 
  ) +
  scale_color_viridis_d(option = "H")
ggsave(
  filename = "tournament_comparison_plot_rydog.png",  # Change file type if needed
  plot = last_plot(),  # Saves the last ggplot created
  dpi = 600,        
  bg = "white",
  units = "in",
  width = 20,
  height = 13.43
)

# Apply normalization to all columns except "indices"
df_close_rydog <- df_long_rydog %>% filter(Method %in% c("Single.Elimination..Reseeded.", "Double.Elimination", "Group.Stage", "Single.Elimination..Set.Structure." ))
windowsFonts(Arial = windowsFont("Arial"))
ggplot(df_close_rydog, aes(x = indices, y = MI, color = Method, group = Method)) +
geom_line(linewidth = 2.5) +           # Add lines
geom_point(size=3) +          # Add points
labs(
title = "Comparison of Tournament Structures with Normal Strengths (Rydog)",
x = "Top i Number of Teams",
y = "Normalized Mutual Information \\ Entropy (log10 scale)"
) +
theme_minimal() +       # Use a clean theme
theme(text = element_text(family = "Arial", size = 32),
  legend.position = "bottom",
  legend.box = "vertical",
  plot.title = element_text(hjust = 0.5, size = 36),
  axis.title = element_text(size = 30),
  axis.text = element_text(size=30)) +
  guides(color = guide_legend(nrow = 5)) +
  scale_y_log10(
    labels = scales::label_number() 
  ) +
  scale_color_viridis_d(option = "H")
ggsave(
  filename = "tournament_comparison_plot_close_rydog.png",  # Change file type if needed
  plot = last_plot(),  # Saves the last ggplot created
  dpi = 600,        
  bg = "white",
  units = "in",
  width = 20,
  height = 13.43
)
```

>>>>

mut_info <- function(sims, num_teams, rank_weights = NULL, wtype = "symmetric_salience"){
  # Handle missing weights (default to equal weights if not provided)
  if (is.null(rank_weights)) {
    rank_weights <- rep(1, num_teams)
  }

  # Extract rank_hat lists, widen into columns y1..yN
  y_df <- sims %>%
    group_by(simulation) %>%
    summarise(rank_hat = list(rank_hat), .groups = "drop") %>%
    mutate(rank_hat = map(rank_hat, unlist)) %>%
    unnest_wider(rank_hat, names_sep = "_")

  colnames(y_df)[-1] <- paste0("y", 1:num_teams)

  # Probability of each observed permutation
  prob_df <- y_df %>%
    select(starts_with("y")) %>%
    group_by(across(everything())) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(probability = count / sum(count))

@@

  #for (i in 1:num_teams) {
    #x_i <- paste0("x", i)
    #y_i <- paste0("y", i)

    #max_footrule <- sum(rank_weights * abs((1:num_teams) - (num_teams:1)))
    # Loop through each row of prob_df
    for (j in 1:nrow(prob_df)) {
      #x_vals <- as.numeric(prob_df[j, paste0("x", 1:num_teams)])
      y_vals <- as.numeric(prob_df[j, paste0("y", 1:num_teams)])

@@

    }
  #}
  #w <- 1 / (w+1)  # Inverse of squared differences and manually change 0 probabilities to 1 on the inverse
  #w <- w/sum(w)
  # Compute weights dynamically for each row
  weighted_df <- prob_df %>%
    mutate(w = w)

@@

  Raj_dist <- numerator/denominator
  # Return the weighted data frame with mutual information weights
  return(Raj_dist)
}

@@

mi_plots_df <- data.frame(
indices = 1:8,
"Single Elimination (Reseeded)" = mi_se_reseeded,
"Group Stage" = mi_group_stage,
"Double Elimination" = mi_de_norm,
"Single Elimination (Set Structure)" = mi_se_norm,
"Coin Flip Single Elimination" = mi_random,
#"Unif_SE" = mi_se_unif,
"Bad Seeding Structure Single Elimination" = mi_bad,
"Coin Flip Round Robin" = mi_perms,
"Round Robin" = mi_test_rr,
"Worst Possible" = mi_worst,
"Best Possible" = mi_best,
check.names = FALSE
)

@@

## MI Calculation and Plotting

These functions can take a list of dataframes and create mutual information that can then be plotted.

```{r}
compute_mi_for_datasets <- function(dataset_list) {
  # Validate input
  if (!is.list(dataset_list)) {
    stop("dataset_list must be a list of data frames")
  }

  # Get dataset names from the list
  if (!is.null(names(dataset_list)) && all(names(dataset_list) != "")) {
    dataset_names <- names(dataset_list)
  } else {
    dataset_names <- paste0("Dataset_", seq_along(dataset_list))
  }

  # Helper function to add simulation number if needed
  add_simulation_number <- function(df) {
    if (!"simulation" %in% colnames(df)) {
      n_rows <- nrow(df)
      n_sims <- 10000
      rows_per_sim <- round(n_rows / n_sims)

      # Memory efficient: create simulation numbers
      df$simulation <- rep(1:n_sims, each = rows_per_sim)
    }
    return(df)
  }

  # Compute MI for all datasets and store as list column
  mi_list <- lapply(seq_along(dataset_list), function(i) {
    df <- dataset_list[[i]]
    df <- add_simulation_number(df)

    # Compute MI - returns a vector or dataframe
    result <- compute_mut_plots(df)

    # Convert to vector if it's a dataframe
    if (is.data.frame(result)) {
      result <- unlist(result)
    }

    return(result)
  })

  # Return dataframe with list column
  data.frame(
    dataset = dataset_names,
    stringsAsFactors = FALSE
  ) %>%
    mutate(mi = mi_list)
}

# Example usage:
mi_ex <- compute_mi_for_datasets(list(
   singleElimination_normal_noTies_1 = singleElimination_normal_noTies_1,
   singleElimination_normal_noTies_1_badseed = singleElimination_normal_noTies_1_badseed
 ))

plot_mi <- function(mi_results, dataset_names = NULL) {
  # If no dataset names specified, plot all datasets
  if (is.null(dataset_names)) {
    dataset_names <- mi_results$dataset
  }

  # Filter to only requested datasets
  filtered_results <- mi_results %>%
    filter(dataset %in% dataset_names)

  if (nrow(filtered_results) == 0) {
    stop("No matching datasets found in results.")
  }

  # Convert list column to long format for plotting
  plot_df <- filtered_results %>%
    rowwise() %>%
    mutate(
      index = list(1:length(mi)),
      mi_values = list(mi)
    ) %>%
    unnest(cols = c(index, mi_values)) %>%
    ungroup()

  # Create plot
  ggplot(plot_df, aes(x = index, y = mi_values, color = dataset, group = dataset)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    labs(
      title = "Mutual Information Comparison",
      x = "Top K Teams",
      y = "Mutual Information (MI)",
      color = "Dataset"
    ) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    scale_x_continuous(breaks = 1:max(plot_df$index)) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      legend.position = "bottom",
      legend.title = element_text(size = 11, face = "bold"),
      legend.text = element_text(size = 9)
    ) +
    scale_color_brewer(palette = "Set1")
}

plot_mi(mi_ex)
```

### 2025-11-13 simulations.qmd
Recommit changes without large RData file

# Single Elimination

@@

# Double Elimination

@@

# Round Robin

@@

  # Run 10,000 simulations in parallel
  sim_list <- future_lapply(
    X = 1:10,
    FUN = function(rep) {
      simulate_roundrobin_r(
        num_teams = teams,
        distribution = distribution,
        ties = ties,
        r = r
      )
    },
    future.seed = TRUE    # ensures parallel RNG is valid
  )

@@

# Group Stage

```{r}
# Group stage is currently set so only the top 2 in each group advance
group_stage_param_grid <- expand.grid( 
  distribution = c("Normal","Uniform", "Same"), 
  ties = c(FALSE, TRUE), 
  rounds = c(1,2,4),
  group_length = c(4,8),
  third_place = c(FALSE, TRUE)
  )

@@

  # Build key
    # Build the key
  parts <- c(
    "groupStage",
    tolower(distribution),
    ifelse(ties, "ties", "noTies"),
    paste0(rounds),
    paste0(group_length, "groups"),
    if(third_place) "thirdplace"
  )

@@

# Waterfall

# Step Ladder

# Staged Round Robin

# High School Tennis Ladder

>>>>

# Getting Simulations

## Single Elimination

@@

## Double Elimination

@@

## Round Robin

@@

  # Run 10,000 simulations in parallel
  sim_list <- future_lapply(
    X = 1:10000,
    FUN = function(rep) {
      simulate_roundrobin_r(
        num_teams = teams,
        distribution = distribution,
        ties = ties,
        r = r
      )
    },
    future.seed = TRUE    # ensures parallel RNG is valid
  )

@@

## Group Stage

```{r}
# Group stage is currently set so only the top 2 in each group advance
group_stage_param_grid <- expand.grid( 
  distribution = c("Normal","Uniform", "Same"), 
  ties = c(FALSE, TRUE), 
  rounds = c(1,2,4),
  group_length = c(4,8)
  )

@@

  # Build key
    # Build the key
  parts <- c(
    "groupStage",
    tolower(distribution),
    ifelse(ties, "ties", "noTies"),
    paste0(rounds),
    paste0(group_length, "groups")
  )

@@

## Waterfall

## Step Ladder

```{r}
stepLadder_param_grid <- expand.grid( 
  distribution = c("Normal","Uniform", "Same"), 
  series = c(1, 3, 7), 
  bad_seeding_structure = c(FALSE, TRUE) 
  )

stepLadder_run_sim_block <- function(
  teams, distribution, bad_seeding_structure, series, p_block) {

  # Build key
    # Build the key
  parts <- c(
    "stepLadder",
    tolower(distribution),
    paste0(series),
    if (bad_seeding_structure) "badseed"

  )

  # Run 10,000 simulations in parallel
  sim_list <- future_lapply(
    X = 1:10000,
    FUN = function(rep) {
      simulate_step_ladder(
        num_teams = teams,
        distribution = distribution,
        series = series,
        seeding_structure = seeding_structure
      )
    },
    future.seed = TRUE    # ensures parallel RNG is valid
  )

with_progress({
  p_block <- progressor(steps = nrow(stepLadder_param_grid))

  for (i in seq_len(nrow(stepLadder_param_grid))) {
    row <- stepLadder_param_grid[i, ]

    stepLadder_run_sim_block(
      teams        = teams,
      distribution = row$distribution,
      series       = row$series,
      bad_seeding_structure = row$bad_seeding_structure,
      p_block = p_block
    )
  }
})
```

## Staged Round Robin

## High School Tennis Ladder

# Getting Mutual Information

```{r}
for (df in single_elimination_df){
  MI_singleElimination <- compute_mi_for_datasets(df)
}
